---
title: "Player Data for World Cup 2018"
author: "David Kane"
date: June 5, 2018
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(lubridate)
```

The purpose of this essay is to gather and explore data for the 736 players from the 32 teams at the 2018 FIFA World Cup. The tournament runs from June 14 through July 15 and is probably the most popular sporting event in the world.

## Download Player Data

FIFA has made several official player lists available, conveniently (?) changing the format each time. For this exercise, I use the one from early June. The [tabulizer package](https://CRAN.R-project.org/package=tabulizer) makes extracting information from tables included in a PDF document relatively easy. 

```{r, warning=FALSE}
library(tabulizer)
url <- "https://github.com/davidkane9/wc18/raw/master/fifa_player_list_1.pdf"
out <- extract_tables(url, output = "data.frame")
```

We now have a 32 element list, each item a dataframe of information about the 23 players on each team. Let's combine this information into a single tidy tibble.

```{r}
x <- bind_rows(out) %>% 
  as_tibble() %>% 
  rename(team = Team,
         number = X.,
         position = Pos.,
         name = FIFA.Popular.Name,
         birth_date = Birth.Date,
         shirt_name = Shirt.Name,
         club = Club,
         height = Height,
         weight = Weight) %>% 
  mutate(birth_date = dmy(birth_date),
         league = str_sub(club, -4, -2),
         age = interval(birth_date, "2018-06-14") / years(1),
         age_int = as.integer(floor(age))) %>% 
  mutate(team = case_when(
    team == "Korea Republic" ~ "South Korea",
    team == "IR Iran" ~ "Iran",
    TRUE ~ team))
```

```{r}
set.seed(9)
x %>% sample_n(10)
```


Perform some error checking.

```{r}
stopifnot(length(unique(x$team)) == 32)      # There are 32 teams.
stopifnot(all(range(table(x$team)) == 23))   # Each team has 23 players.
stopifnot(x %>% 
            filter(position == "GK") %>% 
            group_by(team) %>% 
            tally() %>% 
            filter(n != 3) %>% 
            nrow() == 0)                     # All teams have 3 goal keepers.
stopifnot(all(x$position %in% 
                c("GK", "DF", "MF", "FW")))  # All players assigned to 1 of 4 positions.

```

Wikipedia includes other player information which might be interesting, especially the number of caps for each player. A "cap" is an appearence in a game for the national team. The rvest package makes scraping data from Wikipedia fairly easy.

```{r}
library(rvest)
html <- read_html("https://en.wikipedia.org/wiki/2018_FIFA_World_Cup_squads")

name <- html_nodes(html, "th a") %>% 
  html_text() %>% 
  as_tibble() %>% 
  filter(! str_detect(value, "^captain$")) %>% 
  slice(1:736)

caps <- html_nodes(html, ".plainrowheaders td:nth-child(5)") %>% 
  html_text()

number <- html_nodes(html, ".plainrowheaders td:nth-child(1)") %>% 
  html_text()

country <- html_nodes(html, ".mw-headline") %>% 
  html_text() %>%
  as_tibble() %>% 
  filter(! str_detect(value, "Group")) %>% 
  slice(1:32)

wiki_data <- tibble(
  number = as.numeric(number),
  name = name$value,
  team = rep(country$value, each = 23),
  caps = as.numeric(caps))

## I prefer the name from Wikipedia. Exercise for the reader: How might we use
## name and shirt_name to confirm that we have lined up the data correctly?
  
y <- left_join(select(x, -name), wiki_data, by = c("team", "number"))
```

For the entire sample of 736 players, there is a clear birth month effect, visible both when looking at calendar months and when aggregating to calendar quarters.

```{r}
x  %>% ggplot(aes(x = month(birth_date))) + geom_histogram(stat = "count")
```

```{r}
x  %>% ggplot(aes(x = quarter(birth_date))) + geom_histogram(stat = "count") 
```

Strangely, the effect is only true for players who will be 25 and over at the start of the World Cup, about 75% of the sample.

```{r}
x %>% ggplot(aes(x = month(birth_date))) + geom_histogram(stat = "count") + facet_wrap(~ cut(age, breaks = c(0, 24, 50)))
```

```{r}
x %>% ggplot(aes(x = quarter(birth_date))) + geom_histogram(stat = "count") + facet_wrap(~ cut(age, breaks = c(0, 24, 50)))
```
Why would that be true? Maybe something to do with players being assigned to the U-23 team? You can see something weird going on for age 24 especially:

```{r}
plot(table(x$age_int))
```

Note how there are way fewer 24 year-olds than we might expect, given the big jump from 23 to 25? Does this connect to the somewhat strange distribution of birth quarters for players that are age 24?

```{r}
x  %>% filter(age_int == 24) %>% ggplot(aes(x = quarter(birth_date))) + geom_histogram(stat = "count") 
```

It almost looks like there are a bunch of 24 year olds who were born in the first and fourth quarter who should be here but are not present for some reason. More likely, this is just random noise. Filtering for other individual ages yields similar non-constant distributions.

Another thing is that I thought that it would be easy to see similar effects to the above if, instead of filtering by age, I were to filter by birth_day. This, of course, generates the same answers as long as you pick a date which matches the ages, which are calculated off the start of the tournament. So, you want to pick birth_day >= "1993-06-15" to get all the players that are 24 or younger.

But then you need to be careful about, say, looking at all the players that are 24 and a half years ago or younger. When you do that, you are adding a bunch more players, all with birth_months in the 6 months before June. So, just changing "1993-06-15" to "1993-01-01" has a dramatic impact on the above bar charts.

I keep thinking that there is something deeper here . . . Will there always be more players with birth months in the 6 months before a tournament --- regardless of when that tournament is --- then in the 6 months after the tournament? Not really but . . .

Another thing of note is that, when running a chi-squared test --- and adding ages from low to high --- you have to get to 28 before it pops significant for birth quarter and to 31 for birth month. Not sure I am doing this right though . . .




